# Day 2

```{r, message = FALSE}
library(tidyverse)
library(here)
library(sp)
```

## Spatial Data

All data occur at some location is space and time. For know we focus on spatial analyses and will later extend this to spatio-temporal analyses. Let $\mathcal{D}$ represent the spatial domain and let $\mathbf{s}$ be a spatial location. In general, we will let $\mathcal{A} \subset \mathcal{D}$ be a subdomain of the spatial region of $\mathbf{D}$. 

**Insert Diagram from class here**


## Types of spatial data

There are three primary types of spatial data that we are going to consider

### Geostatistical data
* Occur everywhere
* continuous support
* examples: temperature, precipitation
    
```{r}
data("NOAA_df_1990", package = "STRbook")
glimpse(NOAA_df_1990)
```
<<<<<<< HEAD


=======


>>>>>>> c4d0e5059b2e45f143a484736aa818f3fe5da6bf
```{r, fig.width=16, fig.height = 9}
## Only plot the states with data
states <- map_data("state") 
states <- states %>%
    subset(!(region %in% c("washington", "oregon", "california", "nevada", "idaho", "utah", "arizona","montana", "wyoming", "colorado", "new mexico"))) 

## generate map
NOAA_df_1990 %>%
    subset(year == 1990 & day == 1 & proc == "Tmax") %>%   
ggplot(aes(x = lon, y = lat, color = z)) +
    geom_point() + 
    facet_wrap(~ month, scales = "free", nrow = 4) +
    geom_polygon(data = states, aes(x = long, y = lat, group = group), 
                 inherit.aes = FALSE, fill = NA, color = "black") +
    scale_color_viridis_c(option = "inferno") +
    ggtitle("Tmax for the first day of each month in 1990")
```    
    

### Areal data
* Occur only over discrete areas
* can be thought of as an integral of a continuous process over a subdomain $\mathcal{A} \in \mathcal{D}$
* examples: cases of a disease by counties, votes in an election by congressional district

```{r}
data("BEA", package = "STRbook")
glimpse(BEA)
data("MOcounties", package = "STRbook")
glimpse(MOcounties)
MOcounties <- left_join(MOcounties, BEA, by = "NAME10")
```

```{r}
ggplot(MOcounties) +
    geom_polygon(aes(x = long,
                     y = lat, # county boundary
                     group = NAME10, # county group
                     fill = log(X1970))) + # log of income
    geom_path(aes(x = long, y = lat, group = NAME10)) +
    scale_fill_viridis_c(limits = c(7.5, 10.2), option = "plasma", name = "log($)") +
    coord_fixed() + 
    ggtitle("1970") +
    # county boundary
    # county group
    # annotations
    xlab("x (m)") +
    ylab("y (m)") +
    theme_bw()
```
    
### Point process data
* The count and location of the data are random
* examples: tornados, lightning strikes
    
```{r}
# uncomment out this line to download the data
# load(url("http://github.com/mgimond/Spatial/raw/master/Data/ppa.RData"))
# save(starbucks, ma, pop, file = here::here("data", "ppa-starbucks.RData"))
load(here::here("data", "ppa-starbucks.RData"))
glimpse(starbucks)
<<<<<<< HEAD
=======
```

```{r}
library(spatstat)
## add the massachusetts polygon
Window(starbucks) <- ma
marks(starbucks) <- NULL
## plot using the plot function from spatstat
plot(starbucks)
>>>>>>> c4d0e5059b2e45f143a484736aa818f3fe5da6bf
```

```{r}
library(spatstat)
## add the massachusetts polygon
Window(starbucks) <- ma
marks(starbucks) <- NULL
## plot using the plot function from spatstat
plot(starbucks)
```


## Animations of spatial data


* Many different file types for spatial data

    * Typically data are in "flat files" like comma-seperated value (CSV) files 
    
    ```{r, eval = FALSE}
    read.csv(here("path", "to", "file.csv"))
    ```
    
    * "shapefiles" which can be read using _rgdal_ or _maptools_ packages
    
    ```{r, eval = FALSE}
    library(rgdal)
    library(maptools)
    ```
    
    * "NetCDF" files cane be read using _ncdf4_ or _RNetCDF_
    
    ```{r, eval = FALSE}
    library(ncdf4)
    library(RNetCDF)
    ```
    
    
## Textbook package

To install the data from the textbook, go to [https://spacetimewithr.org/](https://spacetimewithr.org/) and follow the link to the code.

```{r, eval = FALSE}
# install.packages("devtools")
library(devtools)
install_github("andrewzm/STRbook")
```

Note that this package is relatively large because it contains a decent amount of spatial data. 


```{r}
library(STRbook)
```


## Spatial Visualization



### In Class Activity: 

From [Lab 2.1](https://spacetimewithr.org/code) on the textbook site

```{r, eval = FALSE}
## Wikle, C. K., Zammit-Mangion, A., and Cressie, N. (2019), 
## Spatio-Temporal Statistics with R, Boca Raton, FL: Chapman & Hall/CRC
## Copyright (c) 2019 Wikle, Zammit-Mangion, Cressie
##
## This program is free software; you can redistribute it and/or
## modify it under the terms of the GNU General Public License
## as published by the Free Software Foundation; either version 2
## of the License, or (at your option) any later version.
##
## This program is distributed in the hope that it will be useful,
## but WITHOUT ANY WARRANTY; without even the implied warranty of
## MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
## GNU General Public License for more details.
                                        
library("dplyr")
library("tidyr")
library("STRbook")

## ------------------------------------------------------------------------
locs <- read.table(system.file("extdata", "Stationinfo.dat",
                               package = "STRbook"),
                  col.names = c("id", "lat", "lon"))
Times <- read.table(system.file("extdata", "Times_1990.dat",
                                package = "STRbook"),
                  col.names = c("julian", "year", "month", "day"))
Tmax <- read.table(system.file("extdata", "Tmax_1990.dat",
                               package = "STRbook"))

## ------------------------------------------------------------------------
names(Tmax) <- locs$id

## ------------------------------------------------------------------------
Tmax <- cbind(Times, Tmax)
head(names(Tmax), 10)

## ------------------------------------------------------------------------
Tmax_long <- gather(Tmax, id, z, -julian, -year, -month, -day)
head(Tmax_long)

## ------------------------------------------------------------------------
Tmax_long$id <- as.integer(Tmax_long$id)

## -----------------------------------------------------------
nrow(Tmax_long)
Tmax_long <- filter(Tmax_long, !(z <= -9998))
nrow(Tmax_long)

## ------------------------------------------------------------------------
Tmax_long <- mutate(Tmax_long, proc = "Tmax")
head(Tmax_long)

## ------------------------------------------------------------------------
data(Tmin_long, package = "STRbook")
data(TDP_long, package = "STRbook")
data(Precip_long, package = "STRbook")

## ------------------------------------------------------------------------
NOAA_df_1990 <- rbind(Tmax_long, Tmin_long, TDP_long, Precip_long)

## ------------------------------------------------------------------------
summ <- group_by(NOAA_df_1990, year, proc) %>%  # groupings
        summarise(mean_proc = mean(z))          # operation

## ------------------------------------------------------------------------
NOAA_precip <- filter(NOAA_df_1990, proc == "Precip" & month == 6)
summ <- group_by(NOAA_precip, year, id) %>%
        summarise(days_no_precip = sum(z == 0))
head(summ)

## ------------------------------------------------------------------------
median(summ$days_no_precip)

## -------------------------------------------------------------
grps <- group_by(NOAA_precip, year, id)
summ <- summarise(grps, days_no_precip = sum(z == 0))

## ------------------------------------------------------------------------
NOAA_df_sorted <- arrange(NOAA_df_1990, julian, id)

## ------------------------------------------------------------------------
df1 <- select(NOAA_df_1990, julian, z)
df2 <- select(NOAA_df_1990, -julian)

## ------------------------------------------------------------------------
NOAA_df_1990 <- left_join(NOAA_df_1990, locs, by = "id")

## ------------------------------------------------------------------------
Tmax_long_sel <- select(Tmax_long, julian, id, z)
Tmax_wide <- spread(Tmax_long_sel, id, z)
dim(Tmax_wide)

## ------------------------------------------------------------------------
M <- select(Tmax_wide, -julian) %>% as.matrix()

## -----------------------------------------------------------
library("sp")
library("spacetime")

## ------------------------------------------------------------------------
NOAA_df_1990$date <- with(NOAA_df_1990,
                       paste(year, month, day, sep = "-"))
head(NOAA_df_1990$date, 4)   # show first four elements

## ------------------------------------------------------------------------
NOAA_df_1990$date <- as.Date(NOAA_df_1990$date)
class(NOAA_df_1990$date)

## ------------------------------------------------------------------------
Tmax_long2 <- filter(NOAA_df_1990, proc == "Tmax")
STObj <- stConstruct(x = Tmax_long2,           # data set
                    space = c("lon", "lat"),  # spatial fields
                    time = "date")            # time field
class(STObj)

## ------------------------------------------------------------------------
spat_part <- SpatialPoints(coords = Tmax_long2[, c("lon", "lat")])
temp_part <- Tmax_long2$date
STObj2 <- STIDF(sp = spat_part,
                time = temp_part,
                data = select(Tmax_long2, -date, -lon, -lat))
class(STObj2)

## ------------------------------------------------------------------------
spat_part <- SpatialPoints(coords = locs[, c("lon", "lat")])
temp_part <- with(Times,
                   paste(year, month, day, sep = "-"))
temp_part <- as.Date(temp_part)

## ------------------------------------------------------------------------
Tmax_long3 <- gather(Tmax, id, z, -julian, -year, -month, -day)

## ------------------------------------------------------------------------
Tmax_long3$id <- as.integer(Tmax_long3$id)
Tmax_long3 <- arrange(Tmax_long3,julian,id)

## ------------------------------------------------------------------------
all(unique(Tmax_long3$id) == locs$id)

## ------------------------------------------------------------------------
STObj3 <- STFDF(sp = spat_part,
                time = temp_part,
                data = Tmax_long3)
class(STObj3)

## ------------------------------------------------------------------------
proj4string(STObj3) <- CRS("+proj=longlat +ellps=WGS84")

## ------------------------------------------------------------------------
STObj3$z[STObj3$z == -9999] <- NA
```





