# Spatial Visualization

```{r, message = FALSE}
library(tidyverse)
library(here)
```


* Many different file types for spatial data

    * Typically data are in "flat files" like comma-seperated value (CSV) files 
    
    ```{r, eval = FALSE}
    read.csv(here("path", "to", "file.csv"))
    ```
    
    * "shapefiles" which can be read using _rgdal_ or _maptools_ packages
    
    ```{r, eval = FALSE}
    library(rgdal)
    library(maptools)
    ```
    
    * "NetCDF" files cane be read using _ncdf4_ or _RNetCDF_
    
    ```{r, eval = FALSE}
    library(ncdf4)
    library(RNetCDF)
    ```
    
    
## Spatial data

To install the data from the textbook, go to [https://spacetimewithr.org/](https://spacetimewithr.org/) and follow the link to the code.

```{r, eval = FALSE}
# install.packages("devtools")
library(devtools)
install_github("andrewzm/STRbook")
```

Note that this package is relatively large because it contains a decent amount of spatial data. 


```{r}
library(STRbook)
```






