# Day 3

```{r, message = FALSE}
library(tidyverse)
library(here)
library(sp)
```

## Anouncements

* Course audits

* Show gitHub page for site [https://github.com/jtipton25/STAT-5413](https://github.com/jtipton25/STAT-5413)

    * Show how to download files and data

* Example Gerrymandering

[https://uglygerry.com/](https://uglygerry.com/)

```{r}
library(showtext)
font_add("myfont", here::here("fonts", "Gerry.otf"))
```

```{r fig.showtext=TRUE, fig.width=7, fig.height=7, cache = TRUE}
plot(cars, family = "myfont")
title(
    main = "This Font is made of \n Gerrymandered Political Districts", 
    family = "myfont", cex.main = 1.5
)
```

## Files for spatial data

* Many different file types for spatial data

* Typically data are in "flat files" like comma-seperated value (CSV) files 

```{r, eval = FALSE}
read.csv(here("path", "to", "file.csv"))
```

* "shapefiles" which can be read using _rgdal_ or _maptools_ packages

```{r, eval = FALSE}
library(rgdal)
library(maptools)
```

* "NetCDF" files cane be read using _ncdf4_ or _RNetCDF_

```{r, eval = FALSE}
library(ncdf4)
library(RNetCDF)
```


## Textbook package

To install the data from the textbook, go to [https://spacetimewithr.org/](https://spacetimewithr.org/) and follow the link to the code.

```{r, eval = FALSE}
# install.packages("devtools")
library(devtools)
install_github("andrewzm/STRbook")
```

Note that this package is relatively large because it contains a decent amount of spatial data. 


```{r}
library(STRbook)
```


## Spatial Visualization

### Spatial visualization using _fields_

* Simulate a process with some random locations

```{r}
library(fields)
## longitude and latitude of approximately the center of Arkansas
lon_lat_center <- c(-92.33, 35.00) 

n   <- 1000
## simulate some random locations
lon  <- runif(n, lon_lat_center[1] - 2, lon_lat_center[1] + 2)
lat  <- runif(n, lon_lat_center[2] - 2, lon_lat_center[2] + 2)
y   <- rnorm(n, lat + lon, 0.1)

plot(lon, lat)
```


```{r}
quilt.plot(lon, lat, y, nx = 30, ny = 30)
points(lon, lat, cex = .3)
```

```{r}
quilt.plot(lon, lat, y, nx = 6, ny = 10)
points(lon, lat, cex = .3)
```

* Simulate a process on a regular grid

```{r}
n <- 50^2
## simulate locations on a grid
lon  <- seq(lon_lat_center[1] - 2, lon_lat_center[1] + 2, length = sqrt(n))
lat  <- seq(lon_lat_center[2] - 2, lon_lat_center[2] + 2, length = sqrt(n))
s <- expand.grid(lon, lat)

head(lon)
head(lat)
head(s)

plot(s, cex = 0.3)
```


```{r}
## simulate some fake data with a north/south trend
y <- 120 - 1.5 * s[, 2] + matrix(rnorm(n), sqrt(n), sqrt(n))
image.plot(lon, lat, y, main = "Plot of simulated data")
```

```{r}
contour(lon, lat, y, main = "Contour plot of simulated data")
```


```{r}
image.plot(lon, lat, y, main = "Plot of simulated data")
contour(lon, lat, y, main = "Contour plot of simulated data", add = TRUE,
        nlevels = 10)
```

```{r}
## adding in maps
library(maps)
maps::map("world")
```

```{r}
maps::map("state")
```

```{r}
maps::map("county")
```

```{r}
maps::map("county", "Arkansas")
points(s, cex = 0.3)
```


```{r}
state <- map.where("state", x = s[, 1], y = s[, 2])
head(state)
table(state)

## subset only points in arkansas
dat <- data.frame(
    lon   = s[, 1],
    lat   = s[, 2],
    state = state
)

maps::map("county", "Arkansas")
dat %>%
    subset(state == "arkansas") %>%
    points(cex = 0.3)

# points(subset(dat, state == "arkansas"), cex = 0.3)
```

Plot the simulated data with the county boundaries

```{r}
image.plot(lon, lat, y, main = "Plot of simulated data")
maps::map("county", add = TRUE, lwd = 2)
```


```{r}
## change the aspect ratio
image.plot(lon, lat, y, main = "Plot of simulated data", asp = 1.3)
maps::map("county", add = TRUE, lwd = 2)
```

