# Day 7

```{r, message = FALSE}
library(tidyverse)
library(fields)
library(mvnfast)
library(gstat)
library(sp)
library(MCMCpack)
```

## Common isotropic correlation functions

* [Tobler's law of geography](https://en.wikipedia.org/wiki/Tobler%27s_first_law_of_geography)
    * "Everything is related to everything else, but near things are more related than distant things"
    * These functions follow Tobler's law in that the function decays with distance
    * These functions are proven to be symmetric and positive definite, thus are valid correlation functions

```{r}
## make a function to plot the correlation functions
plot_corr_function <- function(corr_fun, ## notice that this input is a function
                               d = seq(0, 10, length.out = 1000),
                               phi = c(0.1, 0.5, 1, 5, 10),
                               title = NULL,
                               ...) {
    
    C_h <- matrix(0, length(d), length(phi))
    for (i in 1:length(phi)) {
        C_h[, i] <- corr_fun(d, phi[i], ...)
    }
    
    print(
        data.frame(d = d, C_h = c(C_h), phi = factor(rep(phi, each = length(d)))) %>%
            ggplot(aes(x = d, y = C_h, group = phi, color = phi)) +
            geom_line() +
            ylim(c(0, 1)) +
            ggtitle(title)
    )
}
```

#### Exponential correlation function

\begin{align*}
C(d) & = e^{- \frac{d}{\phi} }
\end{align*}

where $\phi$ is the **spatial range** parameter (called the **length-scale** in Gaussian process literature). Note that this function is **not differentiable** at 0.

```{r}
exponential_cor <- function (d, phi, ...) {
    return(exp( - d / phi))
}

plot_corr_function(exponential_cor, title = "Exponential correlation function")
```

Sometimes this is parameterized using the inverse spatial range $\theta = \frac{1}{\phi}$

\begin{align*}
C(d) & = e^{- d\theta}
\end{align*}

#### Squared exponential (Gaussian) correlation functions

\begin{align*}
C(d) & = e^{- (\frac{d}{\phi})^2 }
\end{align*}

Notice that this function **is differentiable** at 0.

```{r}
gaussian_cor <- function (d, phi, ...) {
    return(exp( - (d / phi)^2))
}

plot_corr_function(gaussian_cor, 
                   title = "Gaussian (squared exponential) correlation function")
```



#### Powered exponential correlation functions

\begin{align*}
C(d) & = e^{- (\frac{d}{\phi})^k }
\end{align*}

```{r, out.width = "24%", fig.show = "hold"}
powered_exp_cor <- function (d, phi, k, ...) {
    return(exp( - (d / phi)^k))
}

plot_corr_function(powered_exp_cor, k = 1, 
                   title = "Powered exponential correlation function, k = 1")
plot_corr_function(powered_exp_cor, k = 2, 
                   title = "Powered exponential correlation function, k = 2")
plot_corr_function(powered_exp_cor, k = 3,
                   title = "Powered exponential correlation function, k = 3")
plot_corr_function(powered_exp_cor, k = 4,
                   title = "Powered exponential correlation function, k = 4")
```



For future classes (basis representations) [Visual exploration of Gaussian Processes](https://distill.pub/2019/visual-exploration-gaussian-processes/)

#### Matern correlation functions


\begin{align*}
C(d) & = \frac{2^{1 - \nu}}{\Gamma(\nu)} \left( \sqrt{2 \nu} \frac{d}{\phi} \right)^\nu K_\nu \left( \sqrt{2 \nu} \frac{d}{\phi} \right)
\end{align*}

* $\Gamma(\cdot)$ is the gamma function
* $K_\nu(\cdot)$ is the modified Bessel function of the second kind
* $\phi$ is the range parameter
* $\nu$ is the smoothness parameter

```{r, out.width = "24%", fig.show = "hold", warning = FALSE}
## Use the Matern fucnction from the fields library
plot_corr_function(Matern, range = c(0.1, 0.5, 1, 5, 10), smoothness = 0.15, 
                   title = "Matern correlation function, smoothness = 0.15")
plot_corr_function(Matern, range = c(0.1, 0.5, 1, 5, 10), smoothness = 0.5, 
                   title = "Matern correlation function, smoothness = 0.5")
plot_corr_function(Matern, range = c(0.1, 0.5, 1, 5, 10), smoothness = 5, 
                   title = "Matern correlation function, smoothness = 5")
plot_corr_function(Matern, range = c(0.1, 0.5, 1, 5, 10), smoothness = 50, 
                   title = "Matern correlation function, smoothness = 50")
```

* Special Cases
1) $\nu = \frac{1}{2} \Rightarrow C(d) = e^{- \frac{d}{\phi} }$ is the exponential correlation function
1) $\nu = \frac{3}{2} \Rightarrow C(d) = \left(1 - \frac{\sqrt{3}d}{\phi} \right) e^{- \frac{\sqrt{3}d}{\phi} }$ 
1) $\nu = \frac{5}{2} \Rightarrow C(d) = \left(1 - \frac{\sqrt{5}d}{\phi} + \frac{5 d^2}{3 \phi^2} \right) e^{- \frac{\sqrt{5}d}{\phi} }$ 
1) $\nu \rightarrow \infty \Rightarrow C(d) = e^{- \frac{1}{2} \left( \frac{d}{\phi} \right)^2 }$ is the Gaussian correlation function

* Note: these functions are valid in $\mathcal{R}^2$. There are generalizations of the functions to other geometries (spheres, stream networks, etc.) and higher dimensions ($\mathcal{R}^d$).

* The Gaussian process with a Matern correlation function with parameter $\nu$ is $\lceil \nu \rceil$-1 times differentiable in the mean-square sense

```{r, out.width = "49%", fig.show = "hold"}
d <- seq(0, 10, length = 1000)
nu <- c(1/2, 3/2, 5/2)
C_h <- c(
    Matern(d, range = 1, nu = nu[1]),
    Matern(d, range = 1, nu = nu[2]),
    Matern(d, range = 1, nu = nu[3])
)
dat <- data.frame(
    d = d, 
    C_h = C_h,
    nu = factor(rep(nu, each = length(d)))
)
ggplot(dat, aes(x = d, y = C_h, group = nu, color = nu)) +
    geom_line() + 
    ggtitle("Matern correlation functions")

## simulate some Gaussian processes
dat$y <- c(
    rmvn(
        1, 
        mu    = rep(0, 1000), 
        sigma = Matern(rdist(seq(0, 10, length = 1000)), range = 1, nu = nu[1])
    ),
    rmvn(
        1, 
        mu    = rep(0, 1000), 
        sigma = Matern(rdist(seq(0, 10, length = 1000)), range = 1, nu = nu[2])
        ),
    rmvn(
        1, 
        mu    = rep(0, 1000), 
        sigma = Matern(rdist(seq(0, 10, length = 1000)), range = 1, nu = nu[3])
    )
)
ggplot(dat, aes(x = d, y = y, group = nu, color = nu)) +
    geom_line() + 
    ggtitle("Gaussian process realizations")

```



    
