# Day 11

## Announcements

## Example spatial analysis

```{r, message = FALSE}
library(tidyverse)
library(fields)
library(geoR)
library(STRbook)
library(maps)
library(nlme)
```

## Spatial model fitting

We will use the _NOAA_df_1990_ _data.frame_ from the _STRbook_ package. We will 

```{r}
data("NOAA_df_1990")
## add a factor variable for left_join
NOAA_df_1990$id_factor <- factor(NOAA_df_1990$id)
dat <- NOAA_df_1990 %>%
    subset(year == 1990 & month == 7 & proc == "Tmax") %>%
    group_by(id_factor) %>% 
    summarize(mean_Tmax = mean(z)) #

## add back in the lat/lon variables
dat <- NOAA_df_1990 %>%
    subset(year == 1990 & month == 7 & proc == "Tmax" & day == 1) %>%
    left_join(dat, by = "id_factor") 
```

We can plot the data

```{r}
states <- map_data("state")

dat %>%
    ggplot(aes(x = lon, y = lat, color = mean_Tmax)) +
    geom_point() +
    geom_polygon(data = states, aes(x = long, y = lat, group = group), 
                          colour = "black", fill = NA) +
    scale_color_viridis_c(option = "plasma") +
    ggtitle("Average July max Temperature in 1990")
```

Subset to just the region of interest

```{r}
states <- map_data("state")

dat %>%
    ggplot(aes(x = lon, y = lat, color = mean_Tmax)) +
    geom_point() +
    scale_color_viridis_c(option = "plasma") +
    ggtitle("Average July max Temperature in 1990") +
    geom_polygon(data = states, aes(x = long, y = lat, group = group), 
                 colour = "black", fill = NA, inherit.aes = FALSE) +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)

```


```{r}
dat %>%
    ggplot(aes(x = mean_Tmax)) +
    geom_histogram() +
    ggtitle("Average July max Temperature in 1990")
```

* The histogram looks non-normal. Is this a problem?
    * No. What matters are the residuals
    
```{r}
hist(resid(lm(dat$mean_Tmax ~ dat$lat)))
```


### Fit the maximum likelihood estimate using the _geoR_ package

* Fit the maximum likelihood estimate using an intercept only model
) using the _geoR_ package

```{r}
fit <- geoR::likfit(
    data         = dat$mean_Tmax,
    coords       = cbind(dat$lat, dat$lon),
    cov.model    = "exponential",
    ini.cov.pars = c(var(dat$mean_Tmax), 1) 
)

summary(fit)
AIC(fit)
fit$BIC
```

* Plot the estimated spatial correlation function

Model: $\mathbf{y} \sim N(\mu \mathbf{1}, \sigma^2 \mathbf{I} + \tau^2 \mathbf{R}(\phi))$

```{r}
sigma2 <- fit$nugget
tau2   <- fit$sigmasq
phi    <- fit$phi


d   <- seq(0, 150, length.out = 1000)
cov <- ifelse(d == 0, sigma2, 0) + tau2 * exp( -  d / phi)
cor <- cov / cov[1]
data.frame(
    d   = d[-1], 
    cov = cov[-1]
) %>%
    ggplot(aes(x = d, y = cov)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cov_nugget = cov[1]), 
               aes(x = d_nugget, y = cov_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated covariance function") +
    ylim(c(0, max(cov)))

data.frame(
    d   = d[-1], 
    cor = cor[-1]
) %>%
    ggplot(aes(x = d, y = cor)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cor_nugget = cor[1]), 
               aes(x = d_nugget, y = cor_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated correlation function") + 
    ylim(c(0, max(cor)))
```

* Fit the model with coefficients

```{r}
fit_coef <- geoR::likfit(
    data         = dat$mean_Tmax,
    trend        = ~ dat$lat + dat$lon,
    coords       = cbind(dat$lat, dat$lon),
    cov.model    = "exponential",
    ini.cov.pars = c(var(dat$mean_Tmax), 1) 
)

summary(fit_coef)
## look at the AIC
AIC(fit)
AIC(fit_coef)

## look at the BIC
fit$BIC
fit_coef$BIC
```

* Which model is better fitting the data?

* Plot the estimated spatial correlation function

Model: $\mathbf{y} \sim N(\mu \mathbf{1}, \sigma^2 \mathbf{I} + \tau^2 \mathbf{R}(\phi))$

```{r}
sigma2 <- fit_coef$nugget
tau2   <- fit_coef$sigmasq
phi    <- fit_coef$phi


d   <- seq(0, 150, length.out = 1000)
cov <- ifelse(d == 0, sigma2, 0) + tau2 * exp( -  d / phi)
cor <- cov / cov[1]
data.frame(
    d   = d[-1], 
    cov = cov[-1]
) %>%
    ggplot(aes(x = d, y = cov)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cov_nugget = cov[1]), 
               aes(x = d_nugget, y = cov_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated covariance function") +
    ylim(c(0, max(cov)))

data.frame(
    d   = d[-1], 
    cor = cor[-1]
) %>%
    ggplot(aes(x = d, y = cor)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cor_nugget = cor[1]), 
               aes(x = d_nugget, y = cor_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated correlation function") + 
    ylim(c(0, max(cor)))
```



* Other models are available -- see _?cov.model_ for details

#### Spatial predictions using _geoR_

```{r}
pred_locations <- expand.grid(
  seq(min(dat$lat), max(dat$lat), length.out = 100),
  seq(min(dat$lon), max(dat$lon), length.out = 100)
)
names(pred_locations) <- c("lat", "lon")
pred_locations <- data.frame(pred_locations)
preds <- krige.conv(
  data = dat$mean_Tmax,
  coords       = cbind(dat$lat, dat$lon),
  locations    = cbind(pred_locations$lat, pred_locations$lon),
  krige = krige.control(
    cov.model    = "exponential",
    beta         = fit_coef$beta,
    cov.pars     = fit_coef$cov.pars,
    nugget       = fit_coef$nugget
  )
)

dat_pred <- data.frame(
  preds = preds$predict,
  var   = preds$krige.var,
  lat   = pred_locations$lat,
  lon   = pred_locations$lon
) 
```


```{r, out.width = "49%", fig.show = "hold"}

ggplot(dat_pred, aes(x = lon, y = lat, fill = preds)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Predicted Average July max Temperature in 1990") +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)

ggplot(dat_pred, aes(x = lon, y = lat, fill = var)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Average July max Temperature prediction Variance in 1990")  +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```

* What explains the "patchiness" of the variance estimates?


```{r}
ggplot(dat_pred, aes(x = lon, y = lat, fill = var)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Average July max Temperature prediction Variance in 1990") +
  geom_point(data = dat, aes(x = lon, y = lat), inherit.aes = FALSE,
             color = "white", size = 0.6) +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```

### Fitting a spatial model using the _nlme_ package

```{r}
fit <- gls(
  mean_Tmax ~ 1,
  data = dat, 
  correlation = corExp(form = ~ lat + lon, nugget = TRUE), 
  method = "ML"
)
summary(fit)
summary(fit)$AIC
summary(fit)$BIC
```

* Plot the estimated covariance/correaltion function

Note: the _gls_ function partitions the variance using a single overall variance $\sigma^2_{gls}$ and a proportion constant $\rho \in (0, 1)$ so that our nugget $\sigma^2 = \rho * \sigma^2_{gls}$ and our partial sill $\tau^2 = (1 - \rho) * \sigma^2_{gls}$.

```{r}
## extract the model parameters from the gls fit
phi    <- exp(fit$model[1]$corStruct[1])
rho    <- 1 / (1 + exp(-fit$model[1]$corStruct[2])) 

sigma2 <- fit$sigma^2 * (rho)
tau2   <- fit$sigma^2 * (1-rho)


d   <- seq(0, 150, length.out = 1000)
cov <- ifelse(d == 0, sigma2, 0) + tau2 * exp( -  d / phi)
cor <- cov / cov[1]
data.frame(
    d   = d[-1], 
    cov = cov[-1]
) %>%
    ggplot(aes(x = d, y = cov)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cov_nugget = cov[1]), 
               aes(x = d_nugget, y = cov_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated covariance function") +
    ylim(c(0, max(cov)))

data.frame(
    d   = d[-1], 
    cor = cor[-1]
) %>%
    ggplot(aes(x = d, y = cor)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cor_nugget = cor[1]), 
               aes(x = d_nugget, y = cor_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated correlation function") + 
    ylim(c(0, max(cor)))
```


We can also fit this model using covariates


```{r}
fit_coef <- gls(
  mean_Tmax ~ lat + lon,
  data = dat, 
  correlation = corExp(form = ~ lat + lon, nugget = TRUE), 
  method = "ML"
)
summary(fit_coef)

summary(fit)$AIC
summary(fit_coef)$AIC

summary(fit)$BIC
summary(fit_coef)$BIC
```

* Plot the estimated covariance/correaltion function

Note: the _gls_ function partitions the variance using a single overall variance $\sigma^2_{gls}$ and a proportion constant $\rho \in (0, 1)$ so that our nugget $sigma^2 = \rho * \sigma^2_{gls}$ and our partial sill $\tau^2 = (1 - \rho) * \sigma^2_{gls}$.

```{r}
## extract the model parameters from the gls fit
phi    <- exp(fit_coef$model[1]$corStruct[1])
rho    <- 1 / (1 + exp(-fit_coef$model[1]$corStruct[2])) 

sigma2 <- fit_coef$sigma^2 * (rho)
tau2   <- fit_coef$sigma^2 * (1-rho)


d   <- seq(0, 150, length.out = 1000)
cov <- ifelse(d == 0, sigma2, 0) + tau2 * exp( -  d / phi)
cor <- cov / cov[1]
data.frame(
    d   = d[-1], 
    cov = cov[-1]
) %>%
    ggplot(aes(x = d, y = cov)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cov_nugget = cov[1]), 
               aes(x = d_nugget, y = cov_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated covariance function") +
    ylim(c(0, max(cov)))

data.frame(
    d   = d[-1], 
    cor = cor[-1]
) %>%
    ggplot(aes(x = d, y = cor)) +
    geom_line() +
    geom_point(data = data.frame(d_nugget = d[1], cor_nugget = cor[1]), 
               aes(x = d_nugget, y = cor_nugget), inherit.aes = FALSE) +
    ggtitle("Estimated correlation function") + 
    ylim(c(0, max(cor)))
```



* Other correlation structures avaiable with _?corClasses_

#### Spatial predictions using _nlme_

```{r}
pred_locations <- expand.grid(
  seq(min(dat$lat), max(dat$lat), length.out = 100),
  seq(min(dat$lon), max(dat$lon), length.out = 100)
)
names(pred_locations) <- c("lat", "lon")
pred_locations <- data.frame(pred_locations)

preds <- predict(
  fit_coef, 
  newdata = pred_locations
)

dat_pred <- data.frame(
  preds = preds,
  lat   = pred_locations$lat,
  lon   = pred_locations$lon
) 

ggplot(dat_pred, aes(x = lon, y = lat, fill = preds)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Predicted Average July max Temperature in 1990") +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```


### model fitting with _autokrige_ function

```{r}
library(automap)
dat_points <- SpatialPointsDataFrame(
  coords = cbind(dat$lon, dat$lat),
  data  = data.frame(
    mean_Tmax = dat$mean_Tmax,
    lat = dat$lat,
    lon = dat$lon
  )
)
pred_points <- SpatialPointsDataFrame(
  coords = cbind(dat_pred$lon, dat_pred$lat),
  data  = data.frame(
    lat = dat_pred$lat,
    lon = dat_pred$lon
  )
)
  
fit <- autoKrige(
  mean_Tmax ~ 1,
  input_data = dat_points, 
  new_data = pred_points
)
summary(fit)
```

```{r}
plot(fit)
```

```{r}
fit_coef <- autoKrige(
  mean_Tmax ~ lat + lon,
  input_data = dat_points, 
  new_data = pred_points
)
summary(fit_coef)
```


```{r}
plot(fit_coef)
```

```{r, out.width = "49%", fig.show = "hold"}
dat_pred <- data.frame(
  preds = fit_coef$krige_output@data$var1.pred,
  var   = fit_coef$krige_output@data$var1.var,
  lat   = pred_locations$lat,
  lon   = pred_locations$lon
) 
ggplot(dat_pred, aes(x = lon, y = lat, fill = preds)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Predicted Average July max Temperature in 1990") +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)


ggplot(dat_pred, aes(x = lon, y = lat, fill = var)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Average July max Temperature prediction Variance in 1990")  +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```


## Kriging

* Given a set of $n$ locations $\mathbf{s}_{obs} = \{ \mathbf{s}_1, \ldots, \mathbf{s}_n \}$, the goal is often to make a prediction at a set of $n_0$ new locations $\mathbf{s}_{unobs} = \{\mathbf{s}_{unobs_1}, \ldots, \mathbf{s}_{unobs_{n_0}} \}$.

* Out of date terms (but still commonly used)
    * Ordinary Kriging -- Kriging using a constant mean function
    * Universal Kriging -- Kriging with an estimated polynomial mean function
    * Not really a useful distinction -- now, all Kriging is universal Kriging    

* [Kriging](https://en.wikipedia.org/wiki/Kriging) is named after Danie G. Krige who did early work on spatial interpolation.


* For now, assume $E(y(\mathbf{s})) = 0 \hspace{1em} \forall \mathbf{s}$ 

* Because our model is a Gaussian process, the joint distribution of the process  $\mathbf{y}_{unobs} = (y(\mathbf{s}_{unobs_1}), \ldots, y(\mathbf{s}_{{unobs_{n_0}}}))'$ at the unobserved locations and the process $\mathbf{y}_{obs} = (y(\mathbf{s}_1), \ldots, y(\mathbf{s}_n))'$ at the observed locations is 

\begin{align*}
\begin{pmatrix} \mathbf{y}_{unobs} \\ \dots \\ \mathbf{y}_{obs} \end{pmatrix} & \sim \operatorname{N} \left( \begin{pmatrix} \mathbf{0}_{unobs} \\ \dots \\ \mathbf{0}_{obs} \end{pmatrix} ,
\begin{pmatrix} \Sigma_{unobs} & \vdots & \boldsymbol{\Sigma}_{unobs,obs} \\ \dots \\ \boldsymbol{\Sigma}_{obs, unobs} & \vdots & \boldsymbol{\Sigma}_{obs} \end{pmatrix}
\right),
\end{align*}

where $\boldsymbol{\Sigma}_{unobs} = \{ Cov \left( y(\mathbf{s}_{unobs_i}),  y(\mathbf{s}_{unobs_j}) \right) \}_{i,j=1}^{n_0}$ is the $n_{unobs} \times n_{unobs}$ covariance matrix that represents the covariance function evaluated at the unobserved locations, $\boldsymbol{\Sigma}_{unobs,obs} = \boldsymbol{\Sigma}_{obs,unobs}' = \{ Cov \left( y(\mathbf{s}_{unobs_i}),  y(\mathbf{s}_{j}) \right) \}_{i=1,\ldots, n_0; j=1, \ldots, n}$ is the $n_{unobs} \times n$ cross-covariance matrix that represents the covariance function evaluated between the unobserved and observed locations, and $\boldsymbol{\Sigma}_{obs} = \{ Cov \left( y(\mathbf{s}_{i}),  y(\mathbf{s}_{j}) \right) \}_{i,j=1}^{n}$ is the $n \times n$ covariance matrix that represents the covariance function evaluated at the observed locations.

* Using the properties of [conditional multivariate normal distributions](https://en.wikipedia.org/wiki/Multivariate_normal_distribution#Conditional_distributions), we have the conditional distribution

$$
\begin{align}
\mathbf{y}_{unobs} | \mathbf{y}_{obs} & \sim \operatorname{N} \left( \boldsymbol{\Sigma}_{unobs,obs} \boldsymbol{\Sigma}_{obs}^{-1} \mathbf{y}_{obs}, \boldsymbol{\Sigma}_{unobs,unobs} - \boldsymbol{\Sigma}_{unobs,obs} \boldsymbol{\Sigma}_{obs}^{-1} \boldsymbol{\Sigma}_{obs,unobs} \right)
\end{align}
$$

* Can plug in the MLE estimate $\boldsymbol{\Sigma}(\hat{\boldsymbol{\theta}})$ to get the estimated distribution (not accounting for parameter uncertainty).

* The Kriging approach gets the same solution by not assuming normality and solving for the best linear unbiased predictor (BLUP)

We generate the predictive maps from the example above using this formula

```{r}
fit <- geoR::likfit(
    data         = dat$mean_Tmax,
    coords       = cbind(dat$lat, dat$lon),
    cov.model    = "exponential",
    ini.cov.pars = c(var(dat$mean_Tmax), 1) 
)
```

```{r}
sigma2 <- fit$nugget
tau2   <- fit$sigmasq
phi    <- fit$phi

## prediction locations
pred_locations <- expand.grid(
  seq(min(dat$lat), max(dat$lat), length.out = 100),
  seq(min(dat$lon), max(dat$lon), length.out = 100)
)
names(pred_locations) <- c("lat", "lon")
```

```{r, error = TRUE, cache = TRUE}
## calculate pairwise distance matrices -- be careful with large data
locs        <- cbind(dat$lat, dat$lon)
D           <- rdist(locs)
D_unobs     <- rdist(pred_locations)
D_unobs_obs <- rdist(pred_locations, locs)

Cov           <- diag(nrow(dat)) * sigma2 + tau2 * exp( - D / phi)
Cov_unobs     <- diag(nrow(pred_locations)) * sigma2 + tau2 * exp( - D_unobs / phi)
Cov_unobs_obs <- tau2 * exp( - D_unobs_obs / phi)

Cov_inv <- solve(Cov)
Cov_inv <- chol2inv(chol(Cov))

## Kriging mean -- note we subtract the mean to get mean 0 then add it back in
pred_mean <- Cov_unobs_obs %*% Cov_inv %*% (dat$mean_Tmax - mean(dat$mean_Tmax)) + mean(dat$mean_Tmax)

pred_mean_fast <- Cov_unobs_obs %*% (Cov_inv %*% (dat$mean_Tmax - mean(dat$mean_Tmax))) + mean(dat$mean_Tmax)

## Kriging variance
pred_var      <- diag(Cov_unobs - Cov_unobs_obs %*% Cov_inv %*% t(Cov_unobs_obs))
pred_var_fast <- diag(Cov_unobs) - rowSums((Cov_unobs_obs %*% Cov_inv) * Cov_unobs_obs)

system.time(diag(Cov_unobs_obs %*% Cov_inv %*% t(Cov_unobs_obs)))
system.time(rowSums((Cov_unobs_obs %*% Cov_inv) * Cov_unobs_obs))

all.equal(pred_var, pred_var_fast)
```


```{r, out.width = "49%", fig.show = "hold"}
dat_pred <- data.frame(
  preds = pred_mean,
  var   = pred_var,
  lat   = pred_locations$lat,
  lon   = pred_locations$lon
) 

ggplot(dat_pred, aes(x = lon, y = lat, fill = preds)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Predicted Average July max Temperature in 1990") +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)

ggplot(dat_pred, aes(x = lon, y = lat, fill = var)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Average July max Temperature prediction variance in 1990")  +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```


### Kriging Equations with mean $\mu = 0$



* **Assumptions -- BLUP**
    * We will assume we are wanting predictions at a single site $\mathbf{s}_0$
    * Without loss of generality, assume that $E(y(\mathbf{s})) = 0$ -- can always subtract the mean and add it back in (see code above)
    * $Cov(\mathbf{y}) = \boldsymbol{\Sigma}$ is the covariance of the data
    * $Cov(y(\mathbf{s}_0), y(\mathbf{s}_i)) = \boldsymbol{\Sigma}_{0i}$
    * The optimal prediction at unobserved location $\mathbf{s}_0$ is a **linear combination of the observed sites** $\Rightarrow$ $\hat{y}(\mathbf{s}_0) = \sum_{i=1}^n w_i y(\mathbf{s}_i)$ for a set of unknown weights $w_i$
    * The optimal predictor is **unbiased** $\Rightarrow$ $E(\hat{y}(\mathbf{s}_0)) = 0$

Therefore, unbiasedness implies that $E(\hat{y}(\mathbf{s}_0)) = \sum_{i=1}^n w_i E(y(\mathbf{s}_i)) = \sum_{i=1}^n w_i 0 = 0$. Because $\mu = 0$, this is trivially easy and we only need to minimize the variance to have the BLUP.

The MSE is 

$$
\begin{align*}
MSE & = E((\hat{y}(\mathbf{s}_0) - y(\mathbf{s}_0))^2) \\
& = E(\hat{y}(\mathbf{s}_0)^2 - 2 \hat{y}(\mathbf{s}_0) y(\mathbf{s}_0) + y(\mathbf{s}_0)^2) \\
& = E(\hat{y}(\mathbf{s}_0)^2) - 2 E(\hat{y}(\mathbf{s}_0) y(\mathbf{s}_0)) + E(y(\mathbf{s}_0)^2) 
\end{align*}
$$

Because $\mu = 0$, we have $E(\hat{y}(\mathbf{s}_0)^2) = Var(\hat{y}(\mathbf{s}_0))$, $E(y(\mathbf{s}_0)^2) = Var(y(\mathbf{s}_0))$, and $E(\hat{y}(\mathbf{s}_0)y(\mathbf{s}_0)) = Cov(\hat{y}(\mathbf{s}_0), y(\mathbf{s}_0))$. Thus, the MSE is 

\begin{align*}
MSE & = E(\hat{y}(\mathbf{s}_0)^2) - 2 E(\hat{y}(\mathbf{s}_0) y(\mathbf{s}_0)) + E(y(\mathbf{s}_0)^2) \\
& = Var(\hat{y}(\mathbf{s}_0)) - 2 Cov(\hat{y}(\mathbf{s}_0), y(\mathbf{s}_0)) + Var(y(\mathbf{s}_0)) \\
& Var \left( \sum_{i=1}^n w_i y(\mathbf{s}_i) \right) - 2 Cov( \sum_{i=1}^n w_i y(\mathbf{s}_i), y(\mathbf{s}_0)) + \Sigma_0 \\
& = \sum_{i=1}^n \sum_{j=1}^n w_i w_j \Sigma_{ij} - 2 \sum_{i=1}^n w_i \Sigma_{0i}
 + \Sigma_{00} \\
& = \mathbf{w}' \boldsymbol{\Sigma} \mathbf{w} - 2 \mathbf{w}' \boldsymbol{\Sigma}_{0} + \Sigma_{00}
\end{align*}
  
To minimize the MSE, we take a derivative with respect to the Kriging weights $\mathbf{w}$ and solve the equation when set equal to 0


\begin{align*}
\frac{\partial}{\partial \mathbf{w}} MSE & = \frac{\partial}{\partial \mathbf{w}} \mathbf{w}' \boldsymbol{\Sigma} \mathbf{w} - 2 \mathbf{w}' \boldsymbol{\Sigma}_{0} + \Sigma_{00} \\
& = 2 \boldsymbol{\Sigma} \mathbf{w} - 2 \boldsymbol{\Sigma}_{0}
\end{align*}

which has solution $\mathbf{w} = \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_0$. 


### Kriging Equations with unknown mean

First, we assume the mean as the form $\mu(\mathbf{s}) = \mathbf{X}(\mathbf{s}) \boldsymbol{\beta}$ and can the regression coefficents can be estimated from the MLE (take derivative of the log-likelhood and set equal to 0 -- [generalized least squares](https://en.wikipedia.org/wiki/Generalized_least_squares)) as $\hat{\boldsymbol{\beta}} = (\mathbf{X}' \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}' \boldsymbol{\Sigma}^{-1} \mathbf{y}$. We also assume that the covariate value $\mathbf{x}(\mathbf{s}_0)$ at the prediction location is known 

The Kriging estimate is 

$$
\begin{align*}
\hat{y}(\mathbf{s}_0) = \hat{\mu}(\mathbf{s}_0) + \boldsymbol{\Sigma}_0 \boldsymbol{\Sigma}^{-1} (y(\mathbf{s}) - \hat{\mu}(\mathbf{s})).
\end{align*}
$$

which has variance 

$$
\begin{align*}
Var(\hat{y}(\mathbf{s}_0)) = \boldsymbol{\Sigma}_{00} - \boldsymbol{\Sigma}_0 \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_0,
\end{align*}
$$

Where $\boldsymbol{\Sigma}_{00}$ is the unconditional variance of the process at location $\mathbf{s}_0$. From this equation, we see that the Kriging variance is smaller than the unconditional variance.

We generate the predictive maps from the example above using the Universal Kriging formula

```{r}
fit_coef <- geoR::likfit(
    data         = dat$mean_Tmax,
    trend        = ~ dat$lat + dat$lon,
    coords       = cbind(dat$lat, dat$lon),
    cov.model    = "exponential",
    ini.cov.pars = c(var(dat$mean_Tmax), 1) 
)
```

```{r}
sigma2 <- fit_coef$nugget
tau2   <- fit_coef$sigmasq
phi    <- fit_coef$phi
beta   <- fit_coef$beta

## prediction locations
pred_locations <- expand.grid(
  seq(min(dat$lat), max(dat$lat), length.out = 100),
  seq(min(dat$lon), max(dat$lon), length.out = 100)
)
names(pred_locations) <- c("lat", "lon")
```

```{r, error = TRUE, cache = TRUE}
## calculate pairwise distance matrices -- be careful with large data
locs        <- cbind(dat$lat, dat$lon)
D           <- rdist(locs)
D_unobs     <- rdist(pred_locations)
D_unobs_obs <- rdist(pred_locations, locs)

Cov           <- diag(nrow(dat)) * sigma2 + tau2 * exp( - D / phi)
Cov_unobs     <- diag(nrow(pred_locations)) * sigma2 + tau2 * exp( - D_unobs / phi)
Cov_unobs_obs <- tau2 * exp( - D_unobs_obs / phi)

Cov_inv <- solve(Cov)

## Kriging mean -- note we subtract the mean to get mean 0 then add it back in
pred_mean <- Cov_unobs_obs %*% Cov_inv %*% (dat$mean_Tmax - cbind(1, locs) %*% beta) + as.matrix(cbind(1, pred_locations)) %*% beta

## Kriging variance -- fast form
pred_var <- diag(Cov_unobs) - rowSums((Cov_unobs_obs %*% Cov_inv) * Cov_unobs_obs)
```


```{r, out.width = "49%", fig.show = "hold"}
dat_pred <- data.frame(
  preds = pred_mean,
  var   = pred_var,
  lat   = pred_locations$lat,
  lon   = pred_locations$lon
) 

ggplot(dat_pred, aes(x = lon, y = lat, fill = preds)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Predicted Average July max Temperature in 1990") +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)

ggplot(dat_pred, aes(x = lon, y = lat, fill = var)) +
  geom_raster() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), 
               colour = "black", fill = NA) +
  scale_fill_viridis_c(option = "plasma") +
  ggtitle("Average July max Temperature prediction variance in 1990")  +
  coord_fixed(xlim = range(dat$lon), ylim = range(dat$lat), ratio = 1.3)
```
