# Day 28

```{r, message = FALSE}
library(tidyverse)
library(fields)
library(mvnfast)
library(patchwork)
library(rstan)
## use recommended rstan settings
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(bayesplot)
```

## Announcements

## Intro to stan

- Needs a c++ compiler toolchain. [See here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) for instructions on how to install the `rstan` library
    - instructions for your OS (Windows, MacOC, Linux, and support for different languages, etc. on the bar to the right)
    
- probabilistic programming

- allows you to focus on modelling rather than algorithms for fitting

- Many alternatives ([NIMBLE](https://r-nimble.org/), [JAGS](http://mcmc-jags.sourceforge.net/), [PyMC3](https://pypi.org/project/pymc3/), [PyMC4 - experimental](https://pypi.org/project/pymc4/))

### Example: Linear regression

```{r}
set.seed(444)
n <- 1000
X <- cbind(1, runif(n))
beta <- c(3, 2)
sigma <- 0.25
y <- as.vector(X %*% beta + rnorm(n, 0, sigma))

data.frame(y = y, x = X[, 2], mu = X %*% beta) %>%
    ggplot(aes(x = x, y = y)) +
    geom_point() +
    geom_line(aes(x = x, y = mu), color = "red")
```

- we need to define a stan model

- create a stan model in a folder named `stan_models` in the Rstudio project folder

- print the output of the model `linear-regression.stan`

```{r}
read_lines(here::here("stan_models", "linear-regression.stan"))
```

- What are the priors implied by this model on $\boldsymbol{\beta}$ and $\sigma$?


- Fitting the model

```{r, cache = TRUE, message = FALSE}
fit <- stan(
    file = here::here("stan_models", "linear-regression.stan"),
    data = list(y = y, n = n, X = X, p = ncol(X)),
    iter = 500
)
```

- examine the output from the MCMC fit

```{r}
## only plot the regression parameters
print(fit, probs = c(0.1, 0.9), pars = c("beta", "sigma"))

## trace plots
mcmc_trace(fit, regex_pars = c("beta", "sigma"))

## area plots of posterior
mcmc_areas(fit, regex_pars = c("beta", "sigma"))

## acf plots
mcmc_acf(fit, regex_pars = c("beta", "sigma"))

## violin plots of posterior
mcmc_violin(fit, regex_pars = c("beta", "sigma"))
```

- compare these estimates to the `lm` estimtes

```{r}
summary(lm(y ~ X - 1))
```

#### Posterior predictive checks

- How do we know if the model is fitting the data well?

- posterior predictive checks examples [here](https://mc-stan.org/bayesplot/reference/PPC-distributions.html) and [here](https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html#defining-y-and-yrep)

\begin{align*}
[\tilde{\mathbf{y}} | \mathbf{y}] & = \int [\tilde{\mathbf{y}} | \boldsymbol{\theta} ] [\boldsymbol{\theta} | \mathbf{y}] \,d\boldsymbol{\theta}
\end{align*}

where the integral over $\boldsymbol{\theta}$ is performed using MCMC sampling.

- Using the `bayesplot` package, we can explre the posterior predictive checks (ppcs) using a variety of functions

```{r}
## extract the posterior predictive samples
y_rep <- rstan::extract(fit, pars = "y_rep")$y_rep

## only plot for 25 randomly selected samples
plot_idx <- sample(1:n, 25)

ppc_data(y, y_rep)
```

```{r, cache = TRUE}
ppc_hist(y[1:5], y_rep[, 1:5])
```

```{r, cache = TRUE}
ppc_intervals(y[plot_idx], y_rep[, plot_idx], x = X[, 2][plot_idx])
```

```{r, cache = TRUE}
ppc_ribbon(y[plot_idx], y_rep[, plot_idx], x = X[, 2][plot_idx])
```

```{r, cache = TRUE}
ppc_dens_overlay(y, y_rep)
```

```{r, cache = TRUE}
ppc_ecdf_overlay(y, y_rep)
```

#### Optimization in stan

- We can also fit the model using optimization:
    - max posterior mode, also known as max a posteriori (*MAP*).
    - max penalized likelihood (*MLE*).

```{r}
model <- stan_model(
    file = here::here("stan_models", "linear-regression.stan")
)

mle <- optimizing(model, data = list(y = y, n = n, X = X, p = ncol(X)))

print(mle$par[1:3], digits = 4)
```

### Example: Spatial model in stan

```{r}
## Make some data
set.seed(1)
n <- 100
coords <- cbind(runif(n, 0, 1), runif(n, 0, 1))

X <- cbind(1, rnorm(n))

beta <- as.matrix(c(1, 5))

sigma2 <- 1
tau2 <- 5
phi <- 3 / 0.5

D <- as.matrix(rdist(coords))
R <- exp(- phi * D)
eta <- c(rmvn(1, rep(0,n), tau2 * R))
y <- rnorm(n, X %*% beta + eta, sqrt(sigma2))
```

- Fit the stan model
    - this can be a little slow for fitting relatively small data
    - however, stan uses Hamiltonian Monte Carlo (HMC) which is highly efficient
    - often can get by with much shorter MCMC chains (500-1000 samples -- monitor the effective sample size)
    

```{r, cache = TRUE, message = FALSE}
fit <- stan(
    file = here::here("stan_models", "spatial-regression.stan"),
    data = list(y = y, n = n, X = X, p = ncol(X), coords = coords),
    iter = 1000
) 
```

- examine the output from the MCMC fit

```{r}
## only plot the regression parameters
print(fit, probs = c(0.1, 0.9), pars = c("beta", "sigma2", "tau2", "phi"))

## trace plots
mcmc_trace(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))

## area plots of posterior
mcmc_areas(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))

## acf plots
mcmc_acf(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))
```


### Example: Predictive process model in stan

- use the same data from the previous example

- Fit the stan model
    - this can be a little slow for fitting relatively small data
    - however, stan uses Hamiltonian Monte Carlo (HMC) which is highly efficient
    - often can get by with much shorter MCMC chains (500-1000 samples -- monitor the effective sample size)
    
```{r}
n_knots <- 5^2
knots <- expand.grid(
    seq(min(coords[, 1]), max(coords[, 1]), length.out = sqrt(n_knots)),
    seq(min(coords[, 2]), max(coords[, 2]), length.out = sqrt(n_knots))
)
```

```{r, cache = TRUE, message = FALSE}
fit <- stan(
    file = here::here("stan_models", "predictive-process-regression.stan"),
    data = list(y = y, n = n, X = X, p = ncol(X), 
                coords = coords, n_knots = n_knots, knots = knots),
    iter = 1000
)
```

- examine the output from the MCMC fit

```{r}
## only plot the regression parameters
print(fit, probs = c(0.1, 0.9), pars = c("beta", "sigma2", "tau2", "phi"))

## trace plots
mcmc_trace(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))

## area plots of posterior
mcmc_areas(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))

## acf plots
mcmc_acf(fit, regex_pars = c("beta", "sigma2", "tau2", "phi"))
```




## stan Hints and tips

- if there are issues, run one chain using `chains = 1` option in the `stan()` function
- pay careful attention to the error messages -- once you learn to read these they can be very informative
- if you have multiple chains that have widely different run times, that suggests there are issues with your model
    - explore advanced stan diagnostics
    - examples [here](https://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html), [here](https://mc-stan.org/rstan/reference/stan_plot_diagnostics.html), and [here](https://betanalpha.github.io/assets/case_studies/rstan_workflow.html). 
    
- [General stan best pracitices](https://github.com/stan-dev/stan/wiki/Stan-Best-Practices)