# Day 31

```{r}
library(tidyverse)
library(sp)
library(igraph)
library(Matrix)
library(viridis)
library(patchwork)
```
## Announcements


## Change of Spatial Support

- Also called the ecological fallcy

- Often data occur in continuous space, but are summarized in higher order summaries

    - example: lightning strikes over a landscape occur in continuous space, could summarize these to strikes per county.
    
    - example: temperature occurs at every point, global climate models operate on gridcells (spatial averages)
    
**draw picture**

- sometimes data occur over different scales

    - socio-economic and other data are collected by the US Census on the block, block-group, and tract scale
    
    - infection data (coronavirus, flu) is often reported at the county level
    
    - need to reconcile data that occur over different scales (census tract vs. county) and may be potentially mis-aligned

- Example: Gerrymandering

    - [Interactive Gerrymandering map](https://projects.fivethirtyeight.com/redistricting-maps/#Compact)

        - [details on the mapping](https://fivethirtyeight.com/features/hating-gerrymandering-is-easy-fixing-it-is-harder/)
    

## Modeling change of support -- continuous-discrete modeling

- $y(\mathbf{s})$ is data measured at location $\mathbf{s} \in \mathcal{D}$
- $z_i$ is the data measured for region $A_i \in \mathcal{A}$
- The domains $\mathcal{D}$ and $\mathcal{A}$ are the same -- the difference is whether the underlying spatial process is continuous or discrete

**draw picture**

- Example: 
    - $y(\mathbf{s})$ is tree density measured at location $\mathbf{s}$
    - $z_i$ is satellite data at the grid cell $A_i$

Model: Assume a continuous latent process $\eta(\mathbf{s}) \sim GP$ is a Gaussian process and iid error processes $\varepsilon(\mathbf{s}) \sim N(0, \sigma^2_y)$ and $\delta_i \sim N(0, \sigma^2_z)$ , then the joint model for $y$ and $z$ are

\begin{align*}
y(\mathbf{s}) & = \eta(\mathbf{s}) + \varepsilon(\mathbf{s}) \\
z_i & = a + b \int_{A_i} \eta(\mathbf{s}) \,d\mathbf{s} + \delta_i,
\end{align*}

where $a$ and $b$ are bias terms that account for the averaging over the region $A_i$. In practice, the integral $\int_{A_i} \eta(\mathbf{s}) \,d\mathbf{s}$ is difficult to evaluate and requires approximation.

## Modeling change of support -- discrete-discrete modeling

- $y_i$ is data measured for region $A_i \in \mathcal{A} = \{ A_1, \ldots, A_n \}$
- $z_j$ is the data measured for region $B_j \in \mathcal{B} = \{ B_1, \ldots, B_m \}$
- The domains $\mathcal{A}$ and $\mathcal{B}$ are the same -- the difference is partitioning of the overall domain

**draw picture**

- Example: 
    - $y_i$ is tree density measured at location $\mathbf{s}$
    - $z_i$ is satellite data at the grid cell $A_i$

Model: Define a fine partition $\mathcal{D} = \{ D_1, \ldots, D_p \}$ and let $\boldsymbol{\eta} \sim \operatorname{N} \left(\mathbf{0}, \mathbf{Q}^{-1} \right)$ be a p-dimensional CAR process over the domain $\mathcal{D}$ and iid error processes $\varepsilon_i \sim N(0, \sigma^2_y)$ and $\delta_i \sim N(0, \sigma^2_z)$ , then the joint model for $y$ and $z$ are

\begin{align*}
y_i & = \sum_{k = 1}^p \eta_k |D_k \cap A_i| + \varepsilon_i \\
z_j & = \sum_{k = 1}^p \eta_k |D_k \cap B_j|  + \delta_i,
\end{align*}

where $|D_k \cap A_i|$ is the area of the intersection of partition $D_k$ and $A_i$ and $|D_k \cap B_j|$ is the area of the intersection of partition $D_k$ and $B_j$. As long as the size of the fine partition isn't too large, this is pretty straightforward. In matrix notation, this simplifies to

\begin{align*}
\mathbf{y} = (y_1, \ldots, y_n)' & \sim \operatorname{N}\left( \mathbf{W}_y \boldsymbol{\eta}, \sigma^2_y \mathbf{I} \right) \\
\mathbf{z} = (z_1, \ldots, z_m)' & \sim \operatorname{N}\left( \mathbf{W}_z \boldsymbol{\eta}, \sigma^2_z \mathbf{I} \right) \\
\boldsymbol{\eta} \sim \operatorname{N}\left( \mathbf{0}, \tau^2 \mathbf{Q}^{-1} \right),
\end{align*}

where $\mathbf{W}_y$ is an $n \times p$ matrix where the $i$th row is proportional to the area of the domain $A_i$ in each of the fine partitions $\mathbf{D}$, $\mathbf{W}_z$ is an $m \times p$ matrix where the $j$th row is proportional to the area of the domain $B_j$ in each of the fine partitions $\mathbf{D}$, and $\mathbf{Q}$ is a precision matrix for a CAR process on the domain $\mathcal{D}$

```{r}
set.seed(112)
## fine domain D
n_d <- 100^2
locs_d <- expand.grid(
    seq(0, 1, length.out = sqrt(n_d)),
    seq(0, 1, length.out = sqrt(n_d))
)
grid_d <- SpatialGrid(
    GridTopology(c(0, 0), cellsize = c(abs(locs_d[1, 1] - locs_d[2, 1]), abs(locs_d[1, 1] - locs_d[2, 1])), cells.dim = c(sqrt(n_d), sqrt(n_d)))
)

## observation domain A
n_a <- 33^2
locs_a <- expand.grid(
    seq(0, 1, length.out = sqrt(n_a)),
    seq(0, 1, length.out = sqrt(n_a))
)

grid_a <- SpatialGrid(
    GridTopology(c(0, 0), cellsize = c(abs(locs_a[1, 1] - locs_a[2, 1]), abs(locs_a[1, 1] - locs_a[2, 1])), cells.dim = c(sqrt(n_a), sqrt(n_a)))
)

## observation domain B
n_b <- 17^2
locs_b <- expand.grid(
    seq(0, 1, length.out = sqrt(n_b)),
    seq(0, 1, length.out = sqrt(n_b))
)

grid_b <- SpatialGrid(
    GridTopology(c(0, 0), cellsize = c(abs(locs_b[1, 1] - locs_b[2, 1]), abs(locs_b[1, 1] - locs_b[2, 1])), cells.dim = c(sqrt(n_b), sqrt(n_b)))
)

plot(grid_d)
plot(grid_a, add = TRUE, col = "red")
plot(grid_b, add = TRUE, col = "blue")

## simulate the spatial process over D
phi <- 0.99
A  <- make_lattice(length = sqrt(n_d), dim = 2) %>%
    as_adjacency_matrix(sparse = TRUE)
D <- Diagonal(x = colSums(A))
Q <- D - phi * A
## cholesky (this is a sparse matrix that is upper diagonal)
R <- chol(Q)
y_d <- backsolve(R, rnorm(n_d))

## assign the values to the grid
grid_d <- SpatialPointsDataFrame(grid_d, data.frame(y = y_d))

dat_d <- data.frame(
    lon = locs_d[, 1],
    lat = locs_d[, 2],
    y   = y_d
)
p1 <- ggplot(data = dat_d, aes(x = lon, y = lat, fill = y)) +
    geom_raster() +
    scale_fill_viridis() +
    ggtitle("latent spatial process")

## aggregate the spatial process for A
y_a <- aggregate(grid_d[1], grid_a, mean)
dat_a <- data.frame(
    lon = locs_a[, 1],
    lat = locs_a[, 2],
    y   = y_a@data$y
)
## aggregate the spatial process for B
y_b <- aggregate(grid_d[1], grid_b, mean)
dat_b <- data.frame(
    lon = locs_b[, 1],
    lat = locs_b[, 2],
    y   = y_b@data$y
)

p1 <- ggplot(data = dat_d, aes(x = lon, y = lat, fill = y)) +
    geom_raster() +
    scale_fill_viridis() +
    ggtitle("latent spatial process")
p2 <- ggplot(data = dat_a, aes(x = lon, y = lat, fill = y)) +
    geom_raster() +
    scale_fill_viridis() +
    ggtitle("spatial process over A")
p3 <- ggplot(data = dat_b, aes(x = lon, y = lat, fill = y)) +
    geom_raster() +
    scale_fill_viridis() +
    ggtitle("spatial process over B")

p1 / p2 / p3
```

